---
title: "Individual observations"
fig-width: 4
fig-height: 4
resources:
  - data
  - utils.py
webr:
  packages:
    - readr
include-in-header:
  - text: |
      <style>
        .table.dataframe { width: auto; text-align: right; }
      </style>
---

::: {.callout-note title="Learning outcomes & Chapter summary"}

1. **Apply** the visualization grammars in Altair and ggplot
   to create a basic chart via
   `alt.Chart(data).mark_point().encode(x='...', y='...', color='...')`
   and <br> `ggplot(data, aes(x=..., y=..., color=...)) + geom_point()`

2. **Interpret** relationships between individual observations in a scatterplot without assuming causation.

3. **Encode** an additional quantitative variable with color and **use** tooltips to identify observations.

:::

```{python observations-setup}
#| echo: false
#| output: false

import pandas as pd
from utils import show_df
```

## Why bother?

Learning how to visualize each individual observation separately
can be thought of as the most fundamental or naive way to visualize data.
We're not summarizing the data in any way
and we are not adding any additional decorative or guiding graphical elements.
All we're doing is
putting one graphical point in our chart
for every record in our table.
Like we saw in the introductory chapter,
even this simple form of visualization
can lead to insights that numbers alone could never bring.

## Creating our first chart

It's time to start coding!
Let's start by defining some example data to visualize.
The use of data in libraries that implements a grammar of graphics
generally follow the principles of ["tidy"](http://vita.had.co.nz/papers/tidy-data.html) data.
Tidy data is defined as a set of named data *columns*
(also referred to as *variables* or *fields*)
with one feature each
and rows with one observation each.
We will start by,
loading in a small dataset relating to cars
containing historical trends in automotive design, performance, and efficiency.

:::: {.panel-tabset group="language"}

### Altair

```{python}
#| label: load-cars-python
#| output: false
import altair as alt
import pandas as pd


url = 'https://raw.githubusercontent.com/joelostblom/teaching-datasets/main/cars.csv'
cars = pd.read_csv(url)
cars
```

```{python}
#| label: show-cars-python
#| echo: false
show_df(cars)
```

### ggplot

```{r}
#| label: load-cars-r
#| output: false
library(tidyverse)


url <- 'https://raw.githubusercontent.com/joelostblom/teaching-datasets/main/cars.csv'
cars <- read_csv(url)
cars
```

```{r}
#| label: show-cars-r
#| echo: false
source("utils.R")
show_df(cars)
```
::::

Looking at the table above,
we might start to generate questions about our data.
Perhaps the cars with the best fuel efficiency
(the ones that travel the most `Miles_per_gallon` of fuel)
are also the cars with the least powerful engines (lowest `Horsepower`)?
This sounds reasonable since a more powerful engine
generally uses more fuel.
Let's see if we can answer this question
with a visualization.
To see how these two variables
relate to each other,
we can map them to two visual channels in the chart,
such as `x` and `y`.

::: {.panel-tabset group="language"}

## Altair

First,
we create a canvas linked to the cars dataframe
via `alt.Chart(cars)`.
Then, we indicate what kind of geometric *mark*
we want to use to represent the data.
Here, it is suitable to show the data as points using `mark_point()`:

```{python}
#| label: cars-scatter-alt
alt.Chart(cars).mark_point().encode(
    x='Miles_per_Gallon',
    y='Horsepower'
)
```

## ggplot

First,
we create a canvas linked to the cars dataframe
via `ggplot(cars)`.
Then, we indicate what kind of geometric *mark*
we want to use to represent the data.
Here, it is suitable to show the data as points using `geom_point()`:


```{r}
#| label: cars-scatter-gg
ggplot(cars, aes(x = Miles_per_Gallon, y = Horsepower)) +
    geom_point()
```

:::

::: {.column-margin}

<br><br>
**Different defaults**

You can see that there are some differences in the default choices
made in each library:

Different styles have different advantages,
and they are easy to customize to look however you want,
which we will learn about later.

Including the origin (x=0, y=0),
or zooming in to the data
can both be sensible defaults in different context,
which we will also learn more about in later chapters.

Observations with missing values in either `Miles_per_Gallon` or `Horsepower`
are dropped since there is no way to visualize there.
Some libraries explicitly warn about this
and others don't.

:::

Does the chart answer our question?
That's up to you to explore in the exercise below!

::::: {.callout-default .ex-prompt}

Study the chart above,
what can you say about the relationship
between `Horsepower` and `Miles_per_Gallon`?
If you choose to use statistical terms such as talking
about the correlation between the two variables,
try to express yourself in common English
as if you were explaining what the chart means
to someone of the general public.
Finally,
it's always good to ask yourself
if your interpretation of the chart makes sense,
given your knowledge about how the world works.

:::: {.callout-default .ex-hint}

It can be helpful to start with a particular area of the chart.
For example,
look at the cars with the most powerful engines (highest horsepower),
what is their fuel efficiency?
Now do the same for the cars with the least powerful engines
and compare the two.

::: {.callout-default .ex-solution}

From the chart,
we can see that `Horsepower` and `Miles_per_Gallon`
are "negatively correlated".
In plain English,
this means that cars with more powerful engines (higher Horsepower),
have lower fuel efficiency (lower Miles_per_Gallon).
This makes sense as more powerful engines often consume more gas
than less powerful ones.
Note that we can't make claims about a causal relationship from the chart alone,
but in this case the observed relationship in the chart
supports our mechanistic understanding of the world.

:::

::::

:::::

Our conclusion above makes sense for the data we showed in chart.
But remember that there are other variables in our dataframe,
let's show it again as a reminder.

```{python}
#| label: show-cars-weight
#| echo: false

show_df(cars)
```

What if one of the other columns also influence a car's fuel efficiency?
For example,
it seems reasonable that the weight of the car could impact fuel efficiency,
since more force is needed to propel a larger mass.
To explore `Weight_in_lbs`
together with the other two variables,
we can add more dimensions to this chart,
e.g. specifying that we want to color the points by this new column.
When using the `color` encoding channel,
higher level visualization libraties
will automatically figure out an appropriate color scale to use.
Since `Weight_in_lbs` column contains continuous numerical data,
a continuous gradient will be used here.

::: {.panel-tabset group="language"}

### Altair

Here,
we add the `color` encoding to the chart
so that we can see how all three variables
relate to each other.
Note that the largest value in the `Weight_in_lbs` column
is represented with a dark color
since this has the highest contrast to the light background of the chart.

```{python}
#| label: cars-weight-alt
alt.Chart(cars).mark_point().encode(
    x='Miles_per_Gallon',
    y='Horsepower',
    color='Weight_in_lbs'
)
```

### ggplot

Here,
we add the `color` encoding to the chart
so that we can see how all three variables
relate to each other.
Note that the largest value in the `Weight_in_lbs` column
is represented with a light color
since this has the highest contrast to the dark background of the chart.

```{r}
#| label: cars-weight-gg
ggplot(cars, aes(x = Miles_per_Gallon, y = Horsepower, color = Weight_in_lbs)) +
    geom_point()
```

:::

::: {.column-margin}

**Making space**

Notice the difference in how the legend is added in the two libraries.
In Altair the overall width of the figure was increased
to keep the dimensions of the chart area the same,
whereas in ggplot the width of the figure is kept constant
so the chart area must narrow to accommodate the legend.
In general,
we don't want the chart area to shrink just because we are adding a legend,
so here we would need to manually increase the width of the ggplot chart,
which we will learn about in a future chapter.
<!--TODO should we learn this here? depends on how many figs will need it before it is introduced in a chapter -->
:::

::::: {.callout-default .ex-prompt}

How does a car's weight relate to its fuel efficiency?

:::: {.callout-default .ex-hint}

As with the previous exercise,
you can try to start with a specific area of the chart,
first focusing on finding the fuel efficiency
of the cars with the highest horsepower,
and then compare it to the cars with the lowest horsepower.

::: {.callout-default .ex-solution}

We can see that weight is correlated positively with `Horsepower`
and negatively with `Miles_per_Gallon`.
This aligns with our knowledge about the world:
heavy cars require bigger engines
and more fuel to propel them forward.
However,
remember that we can't say anything about causation from the chart alone,
we can just note that the correlation matches
what we would expect
based on our knowledge about the physics of moving objects.

:::

::::

:::::

As a final note,
you might have reacted to that it was harder
to tell exact values on the color scale
compared to the x and y scales.
This is an important consideration
and we will dive deeper into
which visual channels are the most effective
at presenting information
in a future chapter.

<!-- TODO this section could combine theoretica lknowledge with coding, but the exercises throughout the chapter stays as conceptual exxcreisces to discover and motivate the content --> 

<!-- {{< include ./_extensions/r-wasm/live/_knitr.qmd >}} -->
<!-- ## Exercise mission -->
<!-- Apply what you have learned about individual observations in the following exercises. -->

<!-- With a commitment to sustainability and a vision for a greener future, -->
<!-- Viridia is taking proactive steps to investigate and mitigate its environmental impact. -->
<!-- One significant area of focus is the emissions associated with food consumption and production. -->
<!-- **Minister Reed**, -->
<!-- head of the environment department, -->
<!-- has called on you to step up to help -->
<!-- and tasked with investigating emissions in the city of Viridia. -->


<!-- ::::: {.callout-default .ex-prompt} -->

<!-- To start with your investigation, -->
<!-- you'll need to visualize the emissions data effectively. -->
<!-- This will help unlock insights crucial for the Ministry of Environment. -->
<!-- The data looks like this: -->

<!-- ```{python} -->
<!-- #| label: food-emissions-data -->
<!-- #| echo: false -->

<!-- food_emission = pd.read_csv("data/aggregated__food_data.csv") -->
<!-- show_df(food_emission) -->
<!-- ``` -->

<!-- Remind me why you are including the Total Food Supply in your analysis? -->
<!-- Guide them toward understanding why a complete picture is necessary and why including the Total Food Supply (kg/capita/yr) column matters (alongside the Emissions Kg CO2 / kg product) -->

<!-- ::::: -->

<!-- ::::: {.callout-default .ex-prompt} -->

<!-- It's time to put the Grammar of Graphics to good use. -->
<!-- Create a scatter plot of the food supply and emissions data by following these steps: -->

<!-- 1. Instantiate the chart from the `food_emission` dataframe -->
<!-- 2. Use the point mark to create a point for each observation in the data. -->
<!-- 3. Encode `Total Food Supply (kg/capita/yr)` in the x channel -->
<!--    and `Emissions (Kg CO2 / kg product)` in the y channel. -->

<!-- You can use the code cells below to code your answer: -->

<!-- ::: {.panel-tabset group="language"} -->

<!-- ### Altair -->

<!-- ```{pyodide} -->
<!-- #| label: food-exercise-python-setup -->
<!-- #| setup: true -->
<!-- #| exercise: -->
<!-- #|   - initial_chart_alt -->
<!-- #|   - tooltip_chart_alt -->
<!-- import altair as alt -->
<!-- import pandas as pd -->
<!-- from utils import assert_chart_equal -->

<!-- url = 'data/aggregated__food_data.csv' -->
<!-- food_emission = pd.read_csv(url) -->
<!-- ``` -->

<!-- ```{pyodide} -->
<!-- #| label: food-scatter-python-exercise -->
<!-- #| exercise: initial_chart_alt -->
<!-- alt.Chart(______).mark_point().encode( -->
<!--     x=______, -->
<!--     y=______, -->
<!-- ) -->
<!-- ``` -->

<!-- ```{pyodide} -->
<!-- #| label: food-scatter-python-check -->
<!-- #| exercise: initial_chart_alt -->
<!-- #| check: true -->
<!-- solution = alt.Chart(food_emission).mark_point().encode( -->
<!--     x='Total Food Supply (kg/capita/yr)', -->
<!--     y='Emissions (Kg CO2 / kg product)', -->
<!-- ) -->
<!-- assert_chart_equal(solution, result) -->
<!-- ``` -->

<!-- ### ggplot -->

<!-- ```{webr} -->
<!-- #| label: food-exercise-r-setup -->
<!-- #| setup: true -->
<!-- #| exercise: -->
<!-- #|   - initial_chart_gg -->
<!-- #|   - tooltip_chart_gg -->
<!-- library(ggplot2) -->

<!-- url <- 'data/aggregated__food_data.csv' -->
<!-- food_emission <- readr::read_csv(url, show_col_types=FALSE) -->
<!-- ``` -->

<!-- ```{webr} -->
<!-- #| label: food-scatter-r-exercise -->
<!-- #| exercise: initial_chart_gg -->
<!-- # TODO change column names so that backticks are not required -->
<!-- ggplot(______) + -->
<!--     aes( -->
<!--         x=`______`, -->
<!--         y=`______`, -->
<!--     ) + -->
<!--     geom_point() -->
<!-- ``` -->

<!-- ```{webr} -->
<!-- #| label: food-scatter-r-check -->
<!-- #| exercise: initial_chart_gg -->
<!-- #| check: true -->
<!-- solution <- ggplot(food_emission) + -->
<!--     aes( -->
<!--         x=`Total Food Supply (kg/capita/yr)`, -->
<!--         y=`Emissions (Kg CO2 / kg product)`, -->
<!--     ) + -->
<!--     geom_point() -->
<!-- # TODO fix check -->
<!-- # if (identical(.result, solution)) { -->
<!-- #   list(correct = TRUE, message = "Nice work!") -->
<!-- # } else { -->
<!-- #    list(correct = FALSE, message = "That's incorrect, sorry.") -->
<!-- # } -->
<!-- list(correct = TRUE, message = "This check is not working yet") -->
<!-- ``` -->

<!-- ::: -->

<!-- :::: {.callout-default .ex-hint} -->

<!-- For a reminder of the syntax, -->
<!-- review section @sec-learning-to-speak. -->

<!-- ::: {.callout-default .ex-solution} -->

<!-- ::: {.panel-tabset group="language"} -->

<!-- ### Altair -->

<!-- ```{python} -->
<!-- #| label: food-scatter-alt -->
<!-- alt.Chart(food_emission).mark_point().encode( -->
<!--     x='Total Food Supply (kg/capita/yr)', -->
<!--     y='Emissions (Kg CO2 / kg product)', -->
<!-- ) -->
<!-- ``` -->

<!-- ### ggplot -->

<!-- ```{r} -->
<!-- #| label: load-food-emissions-r -->
<!-- #| echo: false -->
<!-- url <- 'data/aggregated__food_data.csv' -->
<!-- food_emission <- readr::read_csv(url, show_col_types=FALSE) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- #| label: food-scatter-gg -->
<!-- ggplot(food_emission) + -->
<!--     aes( -->
<!--         x=`Total Food Supply (kg/capita/yr)`, -->
<!--         y=`Emissions (Kg CO2 / kg product)`, -->
<!--     ) + -->
<!--     geom_point() -->
<!-- ``` -->

<!-- ::: -->

<!-- ::: -->

<!-- :::: -->

<!-- ::::: -->


<!-- ::::: {.callout-default .ex-prompt} -->

<!-- These are striking results! There are some clear clusters and outliers in the data. -->

<!-- With altair you can add a `tooltip` to your chart. Tooltips are labels that pop up when you hover over a datapoint. You can choose which variable is shown in your tooltip just as you have been using `x`, `y`, and `color`. -->

<!-- Add a `tooltip` to investigate which foods stand out. -->

<!-- ```{pyodide} -->
<!-- #| label: food-tooltip-exercise -->
<!-- #| exercise: tooltip_chart_alt -->
<!-- alt.Chart(food_emission).______.______( -->
<!--     ______ -->
<!--     ______ -->
<!--     tooltip=______, -->
<!-- ) -->
<!-- ``` -->

<!-- :::: {.callout-default .ex-hint} -->
<!-- Take a closer look at the emissions data table - you'll need the correct column name for the tooltip. -->

<!-- ::: {.callout-default .ex-solution} -->
<!-- ```{python} -->
<!-- #| label: food-tooltip-solution -->
<!-- alt.Chart(food_emission).mark_point().encode( -->
<!--     x='Total Food Supply (kg/capita/yr)', -->
<!--     y='Emissions (Kg CO2 / kg product)', -->
<!--     tooltip='Food Item', -->
<!-- ) -->
<!-- ``` -->

<!-- ```{pyodide} -->
<!-- #| label: food-tooltip-check -->
<!-- #| exercise: tooltip_chart_alt -->
<!-- #| check: true -->
<!-- solution = alt.Chart(food_emission).mark_point().encode( -->
<!--     x='Total Food Supply (kg/capita/yr)', -->
<!--     y='Emissions (Kg CO2 / kg product)', -->
<!--     tooltip='Food Item', -->
<!-- ) -->
<!-- assert_chart_equal(solution, result) -->
<!-- ``` -->

<!-- ::: -->

<!-- :::: -->

<!-- ::::: -->
