---
title: "Counts & Proportions"
fig-width: 5
fig-height: 3.75
---

::: {.callout-note title="Learning outcomes & Chapter summary"}

1. **Count** observations in numerical intervals with histograms and in categorical groups with bar charts.

2. **Compare** counts across groups using color, stacking, and faceting.

3. **Choose** histogram bins and **order** categorical bars to make patterns easier to interpret.

:::

In [Summaries](7_summaries.qmd), we explored central values and variability.
Another useful summary is the number of observations in a group or interval.
We will start with histograms of movie runtimes, then explore categorical bar charts
and take a closer look at bin choices and tooltips.

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}


## Counts

```{python}
#| echo: false
#| output: false
import pandas as pd
import altair as alt

meditation = pd.read_csv('data/meditation-stress-sample.csv')
```

```{r}
#| echo: false
#| output: false
library(readr)
library(ggplot2)

meditation = read_csv('data/meditation-stress-sample.csv')
```

Another useful statistics we can compute
is the counts of observations in each group.

::: {.panel-tabset group="language"}

### Altair

We can use the `'count()'` aggregation
to count observations.

```{python}
alt.Chart(meditation).mark_point().encode(
    y='condition',
    x='count()'
)
```

### ggplot

We can use `stat = 'count'`
to count observations.
Since there is only one way of counting
we don't need to specify a function
(in contrast, there are many ways of summarizing:
mean, median, sum, etc).

```{r}
ggplot(meditation, aes(y = condition)) +
    geom_point(stat = 'count')
```

:::

When we are showing an accumulated value like a count or sum,
it can be beneficial to replace our point mark
with a bar
to be able to more clearly compare groups.
Bar charts visualize magnitudes as position and length.
Since the bar is drawn from zero 
it makes the accumulated amount explicit through its length
(you can think of the bar
as representing all the individual observation
stacked on top of each other).

::: {.panel-tabset group="language"}

### Altair

```{python}
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x='count()'
)
```

### ggplot

ggplot discourages people from using bars for summaries,
so the default for `geom_bar` is actually `stat = 'count'`,
which means we can write just `geom_bar()`.

```{r}
#| fig-height: 2
ggplot(meditation, aes(y = condition)) +
    geom_bar()
```

Note that we have reduced the figure height 
so that the bars have a reasonable width.
In this book,
ggplot size adjustments are done
separate from the R code
(i.e. via Quarto chunk options).
There are [various ways
to adjust the figure sizes
in different
environment](https://stackoverflow.com/questions/29587881/increase-plot-size-width-in-ggplot2).

:::

::: {.callout-warning}

### Avoid giant bars

Some libraries have a fixed chart size
and expands any categories
to fill the entire chart
instead of allocating a fixed space per category.
This can lead to giant bars
(and other marks),
particularly when there are fewer categories
as in the figure below.
These are generally difficult to grasp
so the overall figure size should be
adjusted to make the bars more narrow.

```{r}
#| echo: false
meditation |>
    dplyr::filter(condition %in% c('Mindful walking', 'Body scan')) |>
ggplot(aes(y = condition)) +
    geom_bar()
```

:::

::: {.callout-warning}

### Avoid bar charts for mean and median

Remember that the length of the bar suggests
that we started at zero and added something up,
e.g. the count or sum of observations.
Using bars for means have also been observed to
bias people to thinking that the
distribution of points are more likely to lie
within the graphical region
of the bar^[[Bar graphs depicting averages are perceptually misinterpreted:
The within-the-bar bias](https://link.springer.com/article/10.3758/s13423-012-0247-5)]
:::

### Sorting groups

In our chart above,
the groups on the y-axis are sorted in alphabetical order.
To make it easier to see which group has the most observation
and compare groups with similar amounts of observations
and highlight those that have the most and least observations,
we can instead sort by the count.

::: {.panel-tabset group="language"}

### Altair

The `sort()` method takes the name of another
visual channel to sort by,
such as `'x'`, `'color'`, etc.

```{python}
alt.Chart(meditation).mark_bar().encode(
    y=alt.Y('condition').sort('x'),
    x='count()'
)
```

Generally,
it tends to be a bit more visually appealing
to have the largest bar the closest to the value axis as above.
This is not a hard rule however,
and we can reverse the order by passing `'-x'` instead of `'x'`.

### ggplot

The base R function `reorder`
can be used to sort a column
based on another existing column,
e.g. `reorder(column_to_sort, column_to_use_as_sorting_order)`.
It requires an extra step to do it based on a non-existing column,
such as the count.
Here,
we use `add_count` to first create a column (automatically named `n`),
as the sorting key for our x-axis column:

```{r}
#| fig-height: 2
meditation |>
    dplyr::add_count(condition) |>
    ggplot(aes(y = reorder(condition, n))) +
        geom_bar()
```

Generally,
it tends to be a bit more visually appealing
to have the largest bar the closest to the value axis,
which we can achieve with `reorder(condition, -n)`.
This is not a hard rule however.

:::

This advice generally applies too all types of charts,
such as when showing points for means etc.
One exception is 
Unless there is a natural order to the data 

### Stacked bars

Similar to area charts
bar charts can also show multiple groups
as segments stacked on top of each other.
Let's learn how  that more in this exercise:

::: {.callout-default .ex-prompt}


For example,
we might be interested in how the participants of the survey
self-reported their gender.

::: {.panel-tabset group="language"}

### Altair

```{python}
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x='count()',
    color='gender'
)
```

### ggplot

```{r}
#| fig-height: 2
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar()
```

:::

In this chart,
it is easy to see that
all conditions have more female than male respondents.
But if we are trying to figure out which 
Is it also easy to see which had the most male and female respondents?
Why and what could we do to make this difficult comparison easier?

:::: {.callout-default .ex-hint}

The reason that the counts of males are hard to compare between conditions
is the same as for stacked areas.
But the solution is slightly different
since we have more options than just layering the bars on top of each other.

::::: {.callout-default .ex-solution}

It is difficult to compare the counts of male respondents
because the corresponding bar segments do not share a common baseline.
This means that we are left trying to mentally align the bars
to compare their lengths in our mind.
This is much harder than
simply reading a value of the x-axis for the 

We could switch to a point mark,
just as we switched from areas to lines to solve a similar problem,
but there is one more solution available to use for bars
that were not there for areas,
we can put the bars next to each other:

::: {.panel-tabset group="language"}

### Altair

The `yOffset`/`xOffset` encodings
that we used for jittering
can also take a categorical variable
to offset groups next to each other.

```{python}
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x='count()',
    color='gender',
    yOffset='gender'
)
```

### ggplot

The `position` parameter
that we used for jittering
can also take the `'dodge'` value,
which means that the geoms should "dodge"
stacking on top of each other
and be place next to one another instead.

```{r}
#| fig-height: 3
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar(position = 'dodge')
```

Now it is easy to see which condition had the most male participants too!

:::

:::::

::::

:::

::: {.callout-warning}

### Avoid stacked bar charts when exact group comparisons matter

Just like for stacked areas,
it is difficult to compare the bar segments
that do not share a baseline,
since we have to rely on their length
rather than the position of the top of the bars on the same axis.
We should therefore only use stacked bars
if the total (e.g. the total count per condition)
is of primary importance
and the comparisons between the groups
is of secondary importance.

:::

<!-- TODO include 2d counts there? Pin numbers? -->

## Proportions

Proportions (and percentages) are counts that have been normalized
to a total number of observations for a wider group.
They can be effectively visualized with bars
that indicate how big part they are of the whole.

::: {.panel-tabset group="language"}

### Altair

```{python}
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x=alt.X('count()').stack('normalize'),
    color='gender'
)
```

We could change the displayed axis values
from percentages to proportions
by formatting the labels of the axis as 1 decimal floats (`.1f`):

```{python}
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x=alt.X('count()')
    .stack('normalize')
    .axis(labelExpr=alt.expr.format(alt.datum.value, '.1f')),
    color='gender'
)
```

<!-- TODO change to    .axis(format='.1f'), -->

### ggplot

```{r}
#| fig-height: 2
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar(position = 'fill')
```

We could change the displayed axis values
from proportions to percentages
by specifying that the scale labels are percentages:

```{r}
#| fig-height: 2
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar(position = 'fill') +
    scale_x_continuous(labels = scales::label_percent())
```

:::


When we show proportions in a stacked bar chart
with only two groups,
both groups are easy to compare since the second group
(i.e. the male group above)
also share a common baseline at the top of the scale
(i.e. where the proportion is 1 to the right).
A third group would however be difficult to compare 
since it would be a floating segment in the middle
of the bars without a common baseline.

Proportional bars can also be put next to each other

For proportions,
it is also 

Based on these results we can see there were mostly females

### Pie charts

Pie charts use angles and areas to indicate magnitude.
They should generally be avoided
as stacked and offset bar charts
tend to communicate the same insights more effectively.
A minor point in favor of pie charts
is that they can make even proportions (such as 1/4 and 1/2)
more quickly recognizable than a bar chart
and can thus be helpful for more illustrative purposes
when there are max 2-3 groups.

::: {.panel-tabset group="language"}

### Altair

`Theta` is the encoding channel that controls the angle/size
of each slice in the pie.

```{python}
alt.Chart(meditation).mark_arc().encode(
    theta=alt.Theta('count()').stack(True),
    color='gender'
)
```

We could also [set the inner and outer radius](https://altair-viz.github.io/gallery/radial_chart.html)
if we want to create a donut chart
or radial bar chart.

### ggplot

The grammar of graphics defines a pie chart
as a bar chart projected into polar coordinate space,
which ggplot allows us to do via `coord_polar()`.

```{r}
ggplot(meditation, aes(x = factor(1), fill = gender)) +
    geom_bar() +
    coord_polar(theta = 'y')
```

:::

::: {.callout-warning}

### Avoid pie charts other than for simple purposes

It is almost always better to use some form of a bar chart.
When there are many groups in the data,
pie charts can be disorienting.
In the chart below you can see the results from the Swedish election in 2026,
with parties designated
to the block they identified with before the election.
In the bar chart to the left,
you can immediately see that Block 1 got slightly more votes
whereas in the middle pie chart
it is impossible to determine this information.
Even in the simplified rightmost pie chart
that shows only the votes per block,
it is really difficult to see that the angle/area of block 1
is slightly greater than block 2.

```{python}
#| echo: false
data = (
    ['V', 30, 'Block1'],
    ['S', 99, 'Block1'],
    ['MP', 22, 'Block1'],
    ['C', 25, 'Block1'],
    ['L', 19, 'Block2'],
    ['KD', 22, 'Block2'],
    ['M', 70, 'Block2'],
    ['SD', 62, 'Block2'],
)

election = pd.DataFrame(data, columns=["Party", "Mandate", 'Block'])

bars = alt.Chart(election, height=180).mark_bar().encode(
    x=alt.X("Block:N", title=None),
    y=alt.Y("sum(Mandate)", title="Mandate"),
    color=alt.Color("Party:N").legend(None)
)

pie_parties = alt.Chart(election, width=180, height=180).mark_arc().encode(
    theta=alt.Theta("sum(Mandate):Q", title="Mandate"),
    color=alt.Color("Party:N").legend(None),
    # detail=alt.Color("Party:N"),
)

pie_blocks = pie_parties.encode(
    color=alt.Color("Block:N").legend(None),
).properties(
    title='One color per block'
)

alt.hconcat(
    alt.hconcat(
        bars,
        pie_parties,
        title=alt.Title('One color per party', anchor='middle'),
        spacing=30
    ),
    pie_blocks,
    spacing=60
)
```

<!-- TODO include this example on the color page when talking about appropriate inherent colors -->

:::

Another way of thinking of these count charts is that we are viewing distributions of categorical values.
This concept can also be applied to quantitative values as we will see in the next chapter.
