---
title: "Counts & proportions"
fig-width: 5
fig-height: 3.75
---

::: {.callout-note title="Learning outcomes & Chapter summary"}

1. **Count** observations in categorical groups and **visualize** them with points and bars.

2. **Compare** counts across groups using color, stacked bars, and side-by-side bars.

3. **Distinguish** counts from proportions and **identify** the denominator used to calculate a proportion.

4. **Order** categorical bars and **choose** between bar and pie charts to make comparisons easier.

:::

In [Summaries](7_summaries.qmd), we explored central values and variability.
Another useful summary is the number of observations in a group.
We will return to the meditation dataset to compare counts across conditions,
then explore proportions and different ways to show parts of a whole.

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}


## Counts

```{python}
#| label: counts-python-setup
#| echo: false
#| output: false
import pandas as pd
import altair as alt

meditation = pd.read_csv('data/meditation-stress-sample.csv')
```

```{r}
#| label: counts-r-setup
#| echo: false
#| output: false
library(readr)
library(ggplot2)

meditation <- read_csv('data/meditation-stress-sample.csv')
```

We can summarize the number of observations in each group
by computing a count.

::: {.panel-tabset group="language"}

### Altair

We can use the `'count()'` aggregation
to count observations.
By default,
the categories are sorted alphabetically
from top to bottom
on the y-axis.

```{python}
#| label: condition-count-points-alt
alt.Chart(meditation).mark_point().encode(
    x='count()',
    y='condition'
)
```

### ggplot

We can use `stat = 'count'`
to count observations.
This stat computes the counts itself,
so we don't need to supply a summary function such as `mean` or `median`.
By default,
the categories are sorted alphabetically
from bottom to top
on the y-axis.

```{r}
#| label: condition-count-points-gg
ggplot(meditation, aes(y = condition)) +
    geom_point(stat = 'count')
```

:::

When we are showing an accumulated value like a count or sum,
it can be beneficial to replace our point mark
with a bar
to be able to more clearly compare groups.
Bar charts visualize magnitudes as position and length.
Since the bar is drawn from zero,
it makes the accumulated amount explicit through its length
(you can think of the bar
as representing all the individual observations
stacked on top of each other).

::: {.panel-tabset group="language"}

### Altair

```{python}
#| label: condition-count-bars-alt
alt.Chart(meditation).mark_bar().encode(
    x='count()',
    y='condition'
)
```

### ggplot

To discourage the use of bars for summaries,
the default for `geom_bar` is `stat = 'count'`,
which means we can write just `geom_bar()`.

```{r}
#| label: condition-count-bars-gg
#| fig-height: 2
ggplot(meditation, aes(y = condition)) +
    geom_bar()
```

Note that we have reduced the figure height
so that the horizontal bars have a reasonable thickness.
In this book,
ggplot size adjustments are done
separately from the R code
(via Quarto chunk options).
There are [various ways
to adjust the figure sizes
in different
environments](https://stackoverflow.com/questions/29587881/increase-plot-size-width-in-ggplot2).

:::

::: {.callout-warning}

### Avoid giant bars

Some libraries have a fixed chart size
and expand the space allocated to categories
to fill the entire chart
instead of allocating a fixed space per category.
This can lead to giant bars
(and other marks),
particularly when there are few categories,
as in the figure below.
Excessively thick bars can make a chart feel crowded.
Adjusting the overall figure size can make the bars narrower.

```{r}
#| label: oversized-count-bars-gg
#| echo: false
meditation |>
    dplyr::filter(condition %in% c('Mindful walking', 'Body scan')) |>
    ggplot(aes(y = condition)) +
    geom_bar(fill = '#FF0039')
```

:::

::: {.callout-warning}

### Avoid bar charts for means and medians

Remember that the length of the bar suggests
that we started at zero and added something up,
such as a count or sum.
Using bars for means has also been observed to
bias viewers toward thinking that individual observations are more likely to fall
within the graphical region of the bar than beyond it,
even at equal distances from the mean.^[[Bar graphs depicting averages are perceptually misinterpreted:
The within-the-bar bias](https://link.springer.com/article/10.3758/s13423-012-0247-5)]
:::

### Sorting groups

<!-- TODO also mentions that we should sort points of mean, medians etc -->

In our chart above,
the groups on the y-axis are sorted in alphabetical order.
Sorting by count instead makes it easier to compare groups with similar counts
and identify those with the most and fewest observations.

::: {.panel-tabset group="language"}

#### Altair

The `sort()` method takes the name of another
visual channel to sort by,
such as `'x'`, `'color'`, etc.

```{python}
#| label: condition-sorted-bars-alt
alt.Chart(meditation).mark_bar().encode(
    x='count()',
    y=alt.Y('condition').sort('x')
)
```

This places the largest bar at the bottom,
nearest the x-axis.
To place it at the top instead,
pass `'-x'` rather than `'x'`.
Either direction can work,
but generally it is more visually appealing
to have the longest bar,
the closest to the axis.

#### ggplot

The base R function `reorder()` sets the order of categorical levels
based on another column in the dataframe,
for example,
`reorder(column_to_sort, column_to_use_as_sorting_order)`.
To order the categories by their counts,
we first need to calculate those counts
and add them to the dataframe.
Here,
we use `dplyr::add_count()` to create a column automatically named `n`,
then use it to order the categories on the y-axis.

```{r}
#| label: condition-sorted-bars-gg
#| fig-height: 2
meditation |>
    dplyr::add_count(condition) |>
    ggplot(aes(y = reorder(condition, -n))) +
        geom_bar()
```

We used used `-n` to put the longest bar the closest to the axis.
To put the largest count at the top instead, use `reorder(condition, n)`.
Either direction can work;
but generally it is more visually appealing
to have the longest bar,
the closest to the axis.

:::

Ordering categories by their plotted values can also help in other charts,
such as point charts of group means.
However, when categories have a meaningful order,
such as education levels or days of the week,
preserving that order is generally more useful than sorting by value.
Here,
there is no such natural order,
but we show an example of how we would define one.

::: {.panel-tabset group="language"}

### Altair

The custom order is passed to `sort`,
just as when ordering by another visual channel.

```{python}
#| label: condition-custom-sorted-bars-alt
custom_order = ['Control (book chapter)', 'Mindful breathing', 'Body scan', 'Mindful walking', 'Loving kindness']
alt.Chart(meditation).mark_bar().encode(
    x='count()',
    y=alt.Y('condition').sort(custom_order)
)
```

### ggplot

When ordering by another visual channel
we used `reorder` to change the order of the dataframe column.
To define a custom order,
we need to directly manipulate order of the scale
that the categorical variable is plotting on,
in this case the y scale.

```{r}
#| label: condition-custom-sorted-bars-gg
#| fig-height: 2
custom_order <- c('Control (book chapter)', 'Mindful breathing', 'Body scan', 'Mindful walking', 'Loving kindness')
ggplot(meditation, aes(y = condition)) +
    geom_bar() +
    scale_y_discrete(limits = custom_order)
```
:::

<!-- TODO add counts in 2D, either here or in joint distributions. Might fit better there because all the visual techniques will be the same and can contrast heatmap from -->

### Stacked bars

Like area charts,
bar charts can also show multiple groups
as segments stacked on top of each other.
Let's explore this in the following exercise:

::: {.callout-default .ex-prompt}


For example,
we might be interested in how participants' self-reported gender
is distributed across the study conditions.

::: {.panel-tabset group="language"}

#### Altair

```{python}
#| label: condition-gender-stacked-alt
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x='count()',
    color='gender'
)
```

#### ggplot

```{r}
#| label: condition-gender-stacked-gg
#| fig-height: 2
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar()
```

:::

In this chart,
it is easy to see that
all conditions have more female than male respondents.
Is it equally easy to identify which conditions have the most male and female participants?
Why, and how could we make the harder comparison easier?

:::: {.callout-default .ex-hint}

Look at which bar segments start at zero and which start at another segment's endpoint.
The difficulty is the same as for stacked areas.
But the solution is slightly different
since we have more options than just layering the bars on top of each other.

::::: {.callout-default .ex-solution}

It is difficult to compare counts when the corresponding bar segments
do not share a common baseline.
This means that we are left trying to mentally align the bars
to compare their lengths in our mind.
This is much harder than
reading the endpoints against a shared x-axis baseline.

We could switch to a point mark,
just as we switched from areas to lines to solve a similar problem,
but bars also offer a useful alternative:
we can put them next to each other within each condition.

::: {.panel-tabset group="language"}

##### Altair

The `yOffset`/`xOffset` encodings
that we used for jittering
can also take a categorical variable
to offset groups next to each other.

```{python}
#| label: condition-gender-grouped-alt
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x='count()',
    color='gender',
    yOffset='gender'
)
```

##### ggplot

The `position` parameter
that we used for jittering
can also take the `'dodge'` value,
which places bars next to one another instead of stacking them.

```{r}
#| label: condition-gender-grouped-gg
#| fig-height: 3
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar(position = 'dodge')
```

:::

Now both gender categories can be compared across conditions using a shared zero baseline.
Body scan and mindful breathing tie for the most male participants, with 36 each;
the control condition has the most female participants, with 86.

:::::

::::

:::

::: {.callout-warning}

### Avoid stacked bar charts when exact group comparisons matter

As with stacked areas,
it is difficult to compare the bar segments
that do not share a baseline,
since we have to rely on their length
rather than aligned endpoint positions.
Stacked bars work best when the total, such as the count per condition,
is of primary importance and precise comparisons between segments
are of secondary importance.

:::

<!-- TODO include 2d counts there? Pin numbers? -->

## Proportions

A proportion expresses a part relative to a whole.
For categorical counts, we calculate it by dividing the count for a subgroup
by the total count for the relevant group; multiplying by 100 gives a percentage.
Here, each condition is the whole, and the gender categories are its parts.
Normalizing each bar to the same length lets us compare gender proportions
even though the conditions have different numbers of participants.

::: {.panel-tabset group="language"}

### Altair

```{python}
#| label: condition-gender-proportions-alt
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x=alt.X('count()').stack('normalize'),
    color='gender'
)
```

We could change the displayed axis values
from percentages to proportions
by formatting the axis labels as decimals with one digit after the decimal point (`.1f`).
Here, `labelExpr` formats each tick value directly, overriding the default percentage labels:

```{python}
#| label: condition-gender-proportions-format-alt
alt.Chart(meditation).mark_bar().encode(
    y='condition',
    x=alt.X('count()')
    .stack('normalize')
    .axis(labelExpr=alt.expr.format(alt.datum.value, '.1f')),
    color='gender'
)
```

### ggplot

```{r}
#| label: condition-gender-proportions-gg
#| fig-height: 2
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar(position = 'fill')
```

We could change the displayed axis values
from proportions to percentages
by specifying that the scale labels are percentages:

```{r}
#| label: condition-gender-percentages-gg
#| fig-height: 2
ggplot(meditation, aes(y = condition, fill = gender)) +
    geom_bar(position = 'fill') +
    scale_x_continuous(labels = scales::label_percent())
```

:::


With two segments in each normalized bar,
both proportions can be compared against a shared reference:
one segment starts at zero, and the other ends at one (100%).
With three or more segments, those in the middle generally lack an aligned endpoint,
making them harder to compare.
Proportions can also be shown with side-by-side bars sharing a zero baseline.

Always make the denominator clear.
These bars show proportions **within each condition**, not proportions of the whole sample.
Normalizing also hides differences in sample size: the conditions contain between 84 and 113 participants,
but all five normalized bars have the same total length.
Across the complete 500-observation sample, 344 participants reported female gender (68.8%).

### Pie charts

Pie charts use angles and areas to indicate magnitude.
They should generally be avoided
as bar charts with a shared baseline
tend to communicate the same insights more effectively.
A minor point in favor of pie charts
is that familiar fractions, such as one-quarter and one-half,
can be easy to recognize.
They can therefore be useful for illustration and simple comparisons
with just two or (maybe) three groups.

::: {.panel-tabset group="language"}

#### Altair

`theta` is the visual channel that controls the angle
of each slice in the pie.

```{python}
#| label: gender-pie-alt
alt.Chart(meditation).mark_arc().encode(
    theta=alt.Theta('count()').stack(True),
    color='gender'
)
```

We can also [adjust the inner and outer radii](https://altair-viz.github.io/gallery/radial_chart.html).
Setting an inner radius greater than zero creates a donut chart.

#### ggplot

In ggplot, we can construct a pie chart by transforming
a stacked bar into polar coordinates with `coord_polar()`;
a very grammatically coherent way to express a pie chart!

```{r}
#| label: gender-pie-gg
ggplot(meditation, aes(x = factor(1), fill = gender)) +
    geom_bar() +
    coord_polar(theta = 'y')
```

:::

::: {.callout-warning}

### Avoid pie charts other than for simple purposes

It is almost always better to use some form of a bar chart.
When there are many groups in the data,
pie charts can be disorienting.
In the chart below you can see the results from the Swedish election in 2026,
with parties designated
to the block they identified with before the election.
In the bar chart to the left,
you can immediately see that Block 1 got slightly more votes
whereas in the middle pie chart
it is impossible to determine this information.
Even in the simplified rightmost pie chart
that shows only the votes per block,
it is really difficult to see that the angle/area of block 1
is slightly greater than block 2.

```{python}
#| echo: false
data = (
    ['V', 30, 'Block1'],
    ['S', 99, 'Block1'],
    ['MP', 22, 'Block1'],
    ['C', 25, 'Block1'],
    ['L', 19, 'Block2'],
    ['KD', 22, 'Block2'],
    ['M', 70, 'Block2'],
    ['SD', 62, 'Block2'],
)

election = pd.DataFrame(data, columns=["Party", "Mandate", 'Block'])

bars = alt.Chart(election, height=180).mark_bar().encode(
    x=alt.X("Block:N", title=None),
    y=alt.Y("sum(Mandate)", title="Mandate"),
    color=alt.Color("Party:N").legend(None)
)

pie_parties = alt.Chart(election, width=180, height=180).mark_arc().encode(
    theta=alt.Theta("sum(Mandate):Q", title="Mandate"),
    color=alt.Color("Party:N").legend(None),
    # detail=alt.Color("Party:N"),
)

pie_blocks = pie_parties.encode(
    color=alt.Color("Block:N").legend(None),
).properties(
    title='One color per block'
)

alt.hconcat(
    alt.hconcat(
        bars,
        pie_parties,
        title=alt.Title('One color per party', anchor='middle'),
        spacing=30
    ),
    pie_blocks,
    spacing=60
)
```

<!-- TODO include this example on the color page when talking about appropriate inherent colors -->

:::

Another way of thinking of these count charts is that we are viewing distributions of categorical values.
This concept can also be applied to quantitative values, as we will see in [Binned distributions](9_binned-distributions.qmd).
