---
fig-width: 5
fig-height: 3.75
resources:
  - slides-lec3.pdf
number-sections: false
---

# 99 Misc {.unnumbered}

This placeholder chapter collects material awaiting a permanent home.

::: {.callout-note title="Learning outcomes & Chapter summary"}

1. **Customize** axis labels, scales, titles, legends, and the sizes of plot elements.

2. **Construct** sample distributions and compare their histogram and boxplot representations.

:::

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{python}
#| label: misc-python-setup
#| include: false

import altair as alt
import pandas as pd

alt.data_transformers.enable('vegafusion')
gm = pd.read_csv('https://raw.githubusercontent.com/joelostblom/teaching-datasets/main/world-data-gapminder.csv')
gm2018 = gm.query('year == 2018')
```

```{r}
#| label: misc-r-setup
#| include: false

library(tidyverse)
theme_set(theme_gray(base_size = 18))
gm <- read_csv('https://raw.githubusercontent.com/joelostblom/teaching-datasets/main/world-data-gapminder.csv', show_col_types = FALSE)
gm2018 <- gm %>% filter(year == 2018)
```

The plot-configuration examples use the 2018 Gapminder data introduced in [Summaries](7_summaries.qmd).

## Plot configuration

These two plots are included as a resource
to show how you can change the appearance of a plot,
including titles and sizes of plot elements.
We will later cover what are the best practices in these areas,
but for now it is enough to know the mechanics of how you can change these.
Feel free to explore and create visualizations that appeal to you!

::: {.panel-tabset group="language"}

### Altair

```{python}
#| label: chart-styling-alt
alt.Chart(gm2018, title='My plot title').mark_point().encode(         # Change the title of the plot
    alt.X('child_mortality').title('Child mortality'),                # Change the x-axis title
    alt.Y('life_expectancy').scale(zero=False),                       # Change the y-axis scale to not include zero
    alt.Size('population').scale(range=(100, 1000)),                  # Change the range (min, max) of the size scale to enlarge points
    alt.Tooltip('country')                                            # Add country name on hover
).configure_axis(
    labelFontSize=14,                                                 # Change the font size of the axis labels (the numbers)
    titleFontSize=20                                                  # Change the font size of the axis title
).configure_legend(
    titleFontSize=14                                                  # Change the font size of the legend title
).configure_title(
    fontSize=30                                                       # Change the font size of the chart title
)
```

### ggplot

With ggplot,
the functions have slightly different names 
and are all added to the end of the plotting construct
rather than to the different axes functions directly.

```{r}
#| label: chart-styling-gg
#| fig-width: 6
#| fig-height: 4

ggplot(gm2018, aes(x = child_mortality, y = life_expectancy, size = population)) +
    geom_point(shape = 1, color = 'steelblue') +      # Change to hollow blue circles
    scale_size(range = c(2, 12)) +                    # Change the range of the size scale to enlarge points
    ggtitle('My plot title') +                        # Change the title of the plot
    labs(x = 'Child mortality') +                     # Change the x-axis scale to log and the title to 'POP'
    labs(size = 'Population') +                       # Change legend title
    theme_bw() +                                      # Change the theme to black and white, ther is also theme_classic()
    theme(text =  element_text(size = 24))            # Change the text size of all labels
```

:::

## Optional material

This [short article on kernel bandwidth selection in density estimation](https://aakinshin.net/posts/kde-bw/) is insightful and goes beyond the main chapters on distributions.

## Plots from slides

This is how the distribution plots in the [original lecture slides](slides-lec3.pdf) were created.

```{python}
#| label: slide-distribution-examples
from scipy.stats import pareto


dists = pd.concat([
    pd.DataFrame(
        dict(
            A = pareto.rvs(3, size=100, random_state=100).tolist() + [1.05]*25 + [1.1]*35 + [1.9]*15 + [1.8]*25,
            B = (pareto.rvs(3, size=160, random_state=100) * -1 + 2.85).tolist() + [1.75]*40,
            C = pareto.rvs(3, size=200, random_state=100) * -1 + 2.85,
            D = pareto.rvs(3, size=200, random_state=100),
            E = pareto.rvs(3, size=160, random_state=100).tolist() + [1.05] * 40,
        )
    ),
]).query(
    'A < 2 & B > 1 & C > 1 & D < 2 & E < 2'
).melt(
    var_name='dist_type'
)

# Looping due to a bug in concatenation of faceted charts
# This also gives separate images for saving
for dist_name, dist_df in dists.groupby('dist_type'):
    (alt.Chart(dist_df, height=150, title=dist_name).mark_bar().encode(
        x=alt.X('value', bin=alt.Bin(maxbins=30, extent=(1,1.9)), title='', scale=alt.Scale(domain=(1, 2))),
        y='count()',
    ) &  alt.Chart(dist_df, height=40).mark_boxplot(size=30).encode(
        x=alt.X('value', scale=alt.Scale(zero=False, domain=(1, 1.9)), title=''),
        y=alt.Y('dist_type', title='', axis=alt.Axis(labels=False)),
    )).display()
```
